---
title: "Lesette"
author: "Lesette Campbell"
date: "11/9/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
Wet seasons of October - April 
- easily countable 

rainfall during the dry season (july -oct) is recorded 
- prior to sampling 
- it affects the 
- determinant of dry season survival 

Assumes: 
  pop growth r_t*N_t-1
term for removals
no carrying capacity 

Iniital pop size unknown 

growth rate r varies over time which we relate to rainfall 

t are calendar years 

model growtth rate from t -1 to t 

State-space model for the wildebeest pop and observed is as follow 

N_t | N_t-1 ~ normal(r_t*N_t-1 - C_t-1, simga2_N)

annual growth rate  is modelled a log linear fn of rainfall: 
  
  log(r_t) = \beta_r) + \beta_r1 * rain_t

observation model where observations are assumed to be normally distributed condiontal random variables: 
  
  yhat_t|N_t ~ normal(N_t, simga2_y_t)


Priors: 
N1 ~ dunif(0, 2)                 # Initial population size             
sigma_N ~ dunif(0, 1)            # Initial standard deviation for N 
sigma_obs ~ dunif(0, 1)          # Initial standard deviation for Observation   
sigma2_N <- pow(sigma_N, 2)       # Translating it into variation (sd^2)
sigma2_obs <- pow(sigma_obs, 2)   # Translating it into variation (sd^2)
tau_N <- pow(sigma_N, -2)         # Translate to precision (sd^-2)
tau_obs <- pow(sigma_obs, -2)     # Translate to precision (sd^-2)
beta_0 ~ dnorm(0, 0.01)          # Setting initial beta_0
beta_1 ~ dnorm(0, 0.01)          # Setting initial beta_1

```{r}
library(statsecol)
data("wildebeest")
View(wildebeest)
```

1. Prior Specification: 
 
The priors that require specification in our model include the following: 
- the population size for the wildebeest in the first year(1960) $N_1$
- the variation for our state model and observation model which can be found by stochastically defining the standard deviation and transforming it by a power of 2
- the intercept $\beta_{r_0}$ and  slope $\beta_{r_1}$ of growth as a function of rainfall in the log-linear function $log(r_t) = \beta_{r_0} + \beta_{r_1} * Rain_t$. 


A diffuse prior for the initial population size, and the standard deviations for the state/observation model are defined with a uniform distribution contain a wide range of positive values. A negative population size or standard deviation is unrealistic, and a uniform distribution provides us the capability of specifying non-negative parameter values. The upper range limit for each distribution is based upon picking a value that is substantially greater than the largest known value of that parameter in the data. 
The prior for initial population size is defined as a uniform ~ (0,2), with the largest known being $Nhat = 1.686$. The standard deviation for state model and the observation model is defined as uniform distribution from (0, 1), where the largest known value of $sehat = 0.355$. 

The diffuse priors for \beta_{r_0} and \beta_{r_1} is defined as a normal distribution from (0, 100). Defining the $\beta$'s as a normal distribution allows for them to take on negative or positive values. This is appropriate in the case of rainfall because we are unsure if rainfall is has a negative or positive impact on the population growth. 


```{r}
library(statsecol)
library(jagsUI)
library(ggplot2)
library(zoo)
data("wildebeest")

set.seed(5150)

nyrs <- 30

sink("MT5767_2.txt")
cat("
model{
  #Priors  and constraints: 
  N1 ~ dunif(0, 2)                   # Initial population size  
  N.est[1] <- N1              
  sig.N ~ dunif(0, 1)                # Initial standard deviation for N 
  sig2.N <- pow(sig.N, 2)            # Translating it into variation (sd^2)
  tau.N <- pow(sig.N, -2)            # Translate to precision (sd^-2)
  beta.0 ~ dnorm(0, 0.01)            # Setting initial beta.0
  beta.1 ~ dnorm(0, 0.01)            # Setting initial beta.1

  # Likelihood - State process
  for (t in 1:(nyrs-1)){
  log.r[t] <- beta.0 + beta.1 * rain[t]
    log(r[t]) <- log.r[t]
    N.est[t+1] <- r[t] * N.est[t]            #add removal 
  }
   # Likelihood - Observation process
  for(t in obs.pos){
  y[t] ~ dnorm(N.est[t], pow(sig.y[t], -2))
  }
}
",fill = TRUE)
sink()

obs.pos <- c(2, 4, 6, 8, 12, 13, 18, 19, 23, 25, 27, 30)

wildebeest.data <- list(nyrs = nyrs, 
                        y = wildebeest$Nhat, 
                        rain = wildebeest$rain, 
                        obs.pos = obs.pos,
                        sig.y = wildebeest$sehat)

N.filled <- na.locf(wildebeest$Nhat)


wildebeest.inits <- function(){
  list(
    N1 = runif(1, 0, 2),
    N.est = N.filled,
    sig.N = runif(1, 0, 1),
    beta.0 = runif(1, -2, 2),
    beta.1 = runif(1, -2, 2)
  )
}

wildebeest.parms <- c("beta.0", "beta.1", "sig.N", "N1", "sig.y")

nc <- 5
nb <- 1000
ni <- 10000 + nb
nt <- 1

wildebeest.out <- jags(data = wildebeest.data,
                       inits = wildebeest.inits,
                       parameters.to.save = wildebeest.parms,
                       model.file = "MT5767_2.txt",
                       n.chains = nc,
                       n.iter = ni,
                       n.burnin = nb,
                       n.thin = nt)

library(MCMCvis)
MCMCtrace(wildebeest.out, params = wildebeest.parms[1:5], type = "trace", iter = ni, pdf = FALSE)


```

