---
title: "Lesette"
author: "Lesette Campbell"
date: "11/9/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
Wet seasons of October - April 
- easily countable 

rainfall during the dry season (july -oct) is recorded 
- prior to sampling 
- it affects the 
- determinant of dry season survival 

Assumes: 
  pop growth r_t*N_t-1
term for removals
no carrying capacity 

Iniital pop size unknown 

growth rate r varies over time which we relate to rainfall 

t are calendar years 

model growtth rate from t -1 to t 

State-space model for the wildebeest pop and observed is as follow 

N_t | N_t-1 ~ normal(r_t*N_t-1 - C_t-1, simga2_N)

annual growth rate  is modelled a log linear fn of rainfall: 
  
  log(r_t) = \beta_r) + \beta_r1 * rain_t

observation model where observations are assumed to be normally distributed condiontal random variables: 
  
  yhat_t|N_t ~ normal(N_t, simga2_y_t)


Priors: 
  N1 ~ dunif(0, 2)                 # Initial population size             
sigma_N ~ dunif(0, 0.5)          # Initial standard deviation for N 
sigma_obs ~ dunif(0, 1)          # Initial standard deviation for Observation   
sigma2_N <- pow(sigma_N, 2)       # Translating it into variation (sd^2)
sigma2_obs <- pow(sigma_obs, 2)   # Translating it into variation (sd^2)
tau_N <- pow(sigma_N, -2)         # Translate to precision (sd^-2)
tau_obs <- pow(sigma_obs, -2)     # Translate to precision (sd^-2)
beta_0 ~ dnorm(0, 0.01)          # Setting initial beta_0
beta_1 ~ dnorm(0, 0.01)          # Setting initial beta_1

```{r}
library(statsecol)
data("wildebeest")
View(wildebeest)
```

Justification for priors: 
  uniform lets u cut off at the values at 0 for other priors 
beta_0 and beta_1 can be negative so normal with a wide enough range is appropriate 

